\cleardoublepage
\phantomsection
\addcontentsline{toc}{chapter}{Bibliography}
\begin{thebibliography}{99}
\bibitem{Fisher_36}
R.\ A.\ Fisher: `The use of multiple measurements in taxonomic problems', Annual Eugenics, vol.\ 7, part II, pp.\ 179--188, 1936.
%
\bibitem{Fix_51}
E.\ Fix, J.\ L.\ Hodges: `Discriminatory Analysis. Nonparametric Discrimination: Consistency Properties', USAF School of Aviation Medicine, Randolph Field, Texas, 1951.
%
\bibitem{Nadaraya_64}
E.\ A.\ Nadaraya: `On Estimating Regression', Theory of Probability and Its Applications, vol.\ 9, issue 1, pp.\ 141--142, 1964.
%
\bibitem{Watson_64}
G.\ S.\ Watson: `Smooth Regression Analysis', Sankhyā: The Indian Journal of Statistics, Series A, vol.\ 26, no.\ 4, pp.\ 359--372, 1964.
%
\bibitem{Hart_68}
P.\ E.\ Hart: `The Condensed Nearest Neighbor Rule', IEEE Transactions on Information Theory, vol.\ 18, pp.\ 515–516, 1968.
%
\bibitem{Hoerl_70}
A.\ E.\ Hoerl, R.\ W. Kennard: 'Ridge Regression: Biased Estimation for Nonorthogonal Problems', Technometrics, vol.\ 12, no.\ 1, pp.\ 55--67, 1970.
%
\bibitem{Devroye_78}
L.\ P.\ Devroye: `The Uniform Convergence of the Nadaraya-Watson Regression Function Estimate', The Canadian Journal of Statistics / La Revue Canadienne de Statistique, vol.\ 6, no.\ 2, pp.\ 179--191, 1978.
%
\bibitem{Aeberhard_92}
S. Aeberhard, D. Coomans, O. de Vel: `Comparison of Classifiers in High Dimensional Settings', Tech. Rep. no. 92-02, Dept. of Computer Science and Dept. of Mathematics and Statistics, James Cook University of North Queensland, 1992. 
%
\bibitem{Byrd_95}
R.\ H.\ Byrd, P.\ Lu, J.\ Nocedal: `A Limited Memory Algorithm for Bound Constrained Optimization', SIAM Journal on Scientific and Statistical Computing, vol.\ 16, issue 5, pp.\ 1190--1208, 1995.
%
\bibitem{Tibshirani_96}
R.\ Tibshirani: `Regression Shrinkage and Selection via the Lasso', Journal of the Royal Statistical Society, Series B, vol.\ 58, issue 1, pp.\ 267--288, 1996.
%
\bibitem{Ormoneit_99}
D.\ Ormoneit, T.\ Hastie: `Optimal Kernel Shapes for Local Linear Regression', Advances in Neural Information Processing Systems 12 (NIPS), 1999.
%
\bibitem{Hand_01}
D.\ J.\ Hand, R.\ J.\ Till: `A Simple Generalisation of the Area Under the ROC Curve for Multiple Class Classification Problems', Machine Learning, vol.\ 45, pp.\ 171--186, 2001.
%
\bibitem{Haerdle_04}
W.\ Härdle, M.\ Müller, S.\ Sperlich, A.\ Werwatz: `Nonparametric and Semiparametric Models', Springer, 2004.
%
\bibitem{Hall_04}
P.\ Hall, J.\ Racine, Q.\ Li: `Cross-validation and the Estimation of Conditional Probability Densities', Journal of the American Statistical Association, vol.\ 99, issue 468, pp.\ 1015--1026, 2004.
%
\bibitem{Zou_05}
H.\ Zou, T.\ Hastie: `Regularization and variable selection via the elastic net', Journal of the Royal Statistical Society, Series B, vol.\ 37, part 2, pp.\ 301--320, 2005.
%
\bibitem{Andrew_07}
G.\ Andrew, J.\ Gao: `Scalable Training of $L_1$-Regularized Log-Linear Models', Proceedings of the 24 th International Conference on Machine Learning, 2007.
%
\bibitem{Hayfield_08}
T.\ Hayfield, J.\ S.\ Racine: `Nonparametric Econometrics: The np Package', Journal of Statistical Software, vol.\ 27, issue 5, pp.\ 1--32, 2008.\par
%
Software available: \url{https://cran.r-project.org/web/packages/np}
%
\bibitem{Bontemps_09}
C.\ Bontemps, J.\ S.\ Racine, M.\ Simoni: `Nonparametric vs Parametric Binary Choice Models: An Empirical Investigation', TSE Working Paper Series 09-126, Toulouse School of Economics, 2009.
%
\bibitem{Friedman_10}
J.\ Friedman, T.\ Hastie, R.\ Tibshirani: `Regularization Paths for Generalized Linear Models via Coordinate Descent', Journal of Statistical Software, vol.\ 33, no.\ 1, pp.\ 1--22, 2010.\par
%
Software available: \url{https://cran.r-project.org/web/packages/glmnet}
%
\bibitem{McKinney_10}
W.\ McKinney: `Data Structures for Statistical Computing in Python', Proceedings of the 9th Python Science Conference, pp.\ 56--61, 2010.\par
%
Software available: \url{https://pandas.pydata.org}
%
\bibitem{Pedregosa_11}
F.\ Pedregosa, G.\ Varoquaux, A.\ Gramfort, V.\ Michel, B.\ Thirion, O.\ Grisel, M.\ Blondel, P.\ Prettenhofer, R.\ Weiss, V.\ Dubourg, J.\ Vanderplas, A.\ Passos, D.\ Cournapeau, M.\ Brucher, M.\ Perrot, E.\ Duchesnay: `Scikit-learn: Machine Learning in Python', JMLR 12, pp.\ 2825--2830, 2011.\par
%
Software available: \url{https://scikit-learn.org}
%
\bibitem{Chen_16}
T.\ Chen, C.\ Guestrin: `XGBoost: A Scalable Tree Boosting System', Proceedings of the 22nd ACM SIGKDD International Conference on Knowledge Discovery and Data Mining, pp.\ 785--794, 2016.\par
%
Software available: \url{https://xgboost.ai}
%
\bibitem{Lundberg_17}
S.\ M.\ Lundberg, S.-I.\ Lee: `A Unified Approach to Interpreting Model Predictions', Advances in Neural Information Processing Systems 30 (NIPS), 2017.\par
%
Software available: \url{https://shap.readthedocs.io/en/latest/index.html}
%
\bibitem{Lundberg_18}
S.\ M.\ Lundberg, G.\ G.\ Erion, S.-I.\ Lee: `Consistent Individualized Feature Attribution for Tree Ensembles', arXiv:1802.03888, arxiv.org, 2018.
%
\bibitem{Dua_19}
D.\ Dua, C.\ Graff: UCI Machine Learning Repository, Irvine, CA: University of California, School of Information and Computer Science, 2019.\par
%
Data available: \url{https://archive.ics.uci.edu/ml}
%
\bibitem{Harrell_20}
F.\ Harrel: `Classification vs. Prediction', blog article, 2020.\par
%
Available: \url{https://www.fharrell.com/post/classification/}
%
\bibitem{Harris_20}
C.\ R.\ Harris, K.\ J.\ Millman, S.\ J.\ van der Walt, R.\ Gommers, P.\ Virtanen, D.\ Cournapeau, E.\ Wieser, J.\ Taylor, S.\ Berg, N.\ J.\ Smith, R.\ Kern, M.\ Picus, S.\ Hoyer, M.\ H.\ van Kerkwijk, M.\ Brett, A.\ Haldane, J.\ Fern{\'a}ndez del R{\'i}o, M.\ Wiebe, P.\ Peterson, P.\ G{\'e}rard-Marchant, K.\ Sheppard, T.\ Reddy, W.\ Weckesser, H.\ Abbasi, C.\ Gohlke, T.\ E.\ Oliphant: `Array programming with NumPy', Nature, vol.\ 585, pp.\ 357--362, 2020.\par
%
Software available: \url{https://numpy.org/}
%
\bibitem{Izzo_20}
C.\ Izzo, A.\ Lipani, R.\ Okhrati, F.\ Medda: `A Baseline for Shapely Values in MLPs: from Missingness to Neutrality', arXiv:2006.04896, arxiv.org, 2020.
%
\bibitem{Virtanen_20}
P.\ Virtanen, R.\ Gommers, T.\ E.\ Oliphant, M.\ Haberland, T.\ Reddy, D.\ Cournapeau, E.\ Burovski, P.\ Peterson, W.\ Weckesser, J.\ Bright, S.\ J.\ van der Walt, M.\ Brett, J.\ Wilson, K.\ J.\ Millman, N.\ Mayorov, A.\ R.\ J.\ Nelson, E.\ Jones, R.\ Kern, E.\ Larson, C J Carey, {\.I}. Polat, Y.\ Feng, E.\ W. Moore, J.\ VanderPlas, D.\ Laxalde, J.\ Perktold, R.\ Cimrman, I.\ Henriksen, E.\ A.\ Quintero, C.\ R.\ Harris, A.\ M.\ Archibald, A.\ H.\ Ribeiro, F.\ Pedregosa, P.\ van Mulbregt, SciPy 1.0 Contributors: 'SciPy 1.0: Fundamental Algorithms for Scientific Computing in Python', Nature Methods, vol.\ 17, issue 3, pp.\ 261--272, 2020.\par
%
Software available: \url{https://scipy.org/}
%
\bibitem{Leisch_21}
F.\ Leisch, E.\ Dimitriadou: `mlbench: Machine Learning Benchmark Problems', R package, 2021.\par
%
Software available: \url{https://cran.r-project.org/web/packages/mlbench}
\end{thebibliography}
\endinput
